Artificial Intelligence for the Automated Synthesis and Validation of Programs

### 2018-05-29 doc/programming-project/AI-programming.tex
updates added to the AI-programming proposal

This research project aims to investigate a novel approach for software development, that leverages Artificial Intelligence, to reduce the chances of introducing programming errors. The project proposes to address {\bf automatic program synthesis and validation starting from {\em input-output} tests cases, and using {\em AI planning} as a problem solving engine}. Our work on this particular research topic is the recipient of the {\em 2016 distinguished paper award} at {\sc IJCAI}, the main international conference on Artificial Intelligence.
\end{abstract}

@@

General quality indicators of scientific research: {\bf 373 citations} (287 since 2013) and {\bf h-index 11}. {\small\em - Source google scholar}.

@@

Eva Onaindia is a {\em Full Professor} at the Universitat Politècnica of València. She leads the GRPS where she conducts research in Artificial Intelligence, Automated Planning, Causal Model Learning and Recommender Systems.  She has published numerous articles in top-tier specialized conferences such as the International Joint Conference on Artificial Intelligence (IJCAI), Autonomous Agents and Multi Agent Systems (AAMAS), International Conference on Planning and Scheduling (ICAPS) or European Conference on Artificial Intelligence (ECAI), among others; and scientific journals related to topics of Artificial Intelligence, Planning, Argumentation and Recommender Systems such as IEEE Intelligent Systems, Information Sciences, Engineering Applications of Artificial Intelligence, Knowledge and Information Systems or Applied Intelligence, among others.

She has led several National Research Projects as well as served on multiple ICAPS,AAAI, IJCAI and ECAI conferences as a member of the Program Committee and as a senior member. She is Editor-in-Chief of the journal AI Communications, a Review Board member of the journal Applied Intelligence and she will be Program co-chair of ICAPS 2019 (the main international conference on AI planning).

@@

General quality indicators of scientific research: {\bf 1521 citations} (977 since 2013) and {\bf h-index 22}. {\small\em - Source google scholar}.

@@

\item {\bf Test driven development (TDD)}~\cite{beck:TDD:2003} is a popular paradigm for software development that is frequently used in {\it agile methodologies}~\cite{cohen2003agile}. In TDD, test cases are created before the program code is written and they are run against the code during the development, e.g. after a code change via an automated process. Tests alert programmers of bugs before handing the code off to clients (the cost of finding a bug when the code is first written is considerably lower than the cost of detecting and fixing it later). When all tests pass, the program code is considered {\em complete} while when a test fails, it pinpoints a {\em bug} that must be fixed from the program code. Tests cases are a natural form of program specification, programmers often claim {\em 'code that is difficult to test is poorly written'}. Further, writing a thorough set of tests cases forces programmers to think through inputs, outputs, and error conditions of programs. 
\end{itemize}

@@

Closely related to the aims of the project is the work on the synthesis and validation of {\it Finite State Controllers} (FSCs)~\cite{geffner:policies:IJCAI15}. The state-of-the-art algorithms for computing FSCs follow a {\it top-down} approach that interleaves {\it programming} the FSC with validating it~\cite{sergio:aprograming:ijcai16}. To keep the computation of FSCs tractable, they limit the space of possible solutions bounding the maximum size of the FSC. The computation of FSCs includes works that compile this task into another forms of problem solving so they benefit from the last advances on off-the-shelf solvers (e.g. {\em classical planning}~\cite{sergio:aprograming:icaps16}, {\em conformant planning}~\cite{Geffner:FSM:AAAI10}, {\em CSP}~\cite{Infantes:FSC:ECAI2010} or a {\em Prolog program}~\cite{Giacomo:FSM:ICAPS13}).

Finally the generation of programs from examples is a research question also addressed in inductive Logic Programming (ILP)~\cite{muggleton1991inductive,Raedt:relationalML:book2008}. ILP arises from the intersection of {\em Machine Learning} and {\em Logic Programming} and deals with the development of inductive techniques to learn a given target concept, expressed as a  logic program,  from  examples  and  background  knowledge, that are expressed as logic facts.

(Please find the full list of {\em references} at the end of the document)

@@

\subsection{Synthesis and validation of TDD programs as AI planning}
Given a {\em TDD programming} task our approach is modeling and solving it as it were an {\em AI planning} task. Briefly the state variables of this AIP task are {\em fluents} of the kind {\tt var:=value} (that encode the program variables) plus fluents {\tt program(line):=instruction} (that encode the program lines) and fluents {\tt pcounter:=line} (encoding the current program line). The initial/goal states of the AIP task encode the initial/final values of the program variables according to the input/output tests of the TDD programming task. Finally, program instructions are encoded using two kinds of AIP actions:
\begin{enumerate}
\item {\it Programming actions}, that set an instruction on a given program line.
\item {\it Execution actions}, execute the instruction set on a given program line.
\end{enumerate}
We implement this encoding using standard planning languages, such as PDDL~\cite{fox2003pddl2}, so the AIP tasks resulting from our encoding can be solved with off-the-shelf planners, like {\sc FD}~\cite{helmert2006fast}. The programs synthesized with this approach are guaranteed to be bug-free over given sets of {\em input-output} tests cases.

@@

\subsection{Evaluation}
The performance of our approach for the synthesis and validation of TDD programs will be evaluated in two kinds of tasks:
\begin{itemize}
\item {\em Theoretical benchmarks}: Programming tasks for the computation of mathematical/logic series, string manipulation and for the management of classic data structures such as {\em lists}, {\em queues}, {\em stacks}, {\em trees}, {\em heaps}, \ldots
\item {\em Application benchmarks}: Practical programming tasks coming from {\em activity recognition} and {\em recommendation} domains. The GRPS group is currently leading a four-year research project from the Spanish National plan in which {\em activity recognition} and {\em activity recommendation} are applied to different real-world domains such as {\em tourism}, {\em domotics}, {\em traffic control} and {\em robotics}. 
\end{itemize}

@@

\begin{figure}[hbt!]
\begin{ganttchart}[
  hgrid,
  group progress label node/.append style={below=3pt},
  canvas/.append style={label=below:} ]{1}{12} 
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Design (6 months)}{1}{3} \\
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Development (6 months)}{4}{6}\\
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Dissemination (12 months)}{7}{12}
\end{ganttchart}
\caption{\small Workplan for the development of a user-interactive program synthesizer.}
\label{fig:gantt}
\end{figure}

@@

{\small {\bf Deliverable:} Technical report with the specifications of the system design.}

  \item {\bf T2. Development of the system architecture (months 7-12})
    \begin{small}
      \begin{enumerate}
      \item The programming-into-planning compiler ({\em Compiler 1}). This component parses the {\em TDD programming task} and produces an {\em AIP task} encoded in the standard planning language PDDL.
      \item The plan-into-program compiler ({\em Compiler 2}). This component extracts the program code from the solution plan produced by an off-the-shelf AI planner.
      \end{enumerate}
\end{small}

@@

{\small {\bf Deliverable:} Open repository with the source code of the system architecture.}

\item {\bf T3. Experiments and dissemination of results (months 13-24}).
   \begin{small}
      \begin{enumerate}
      \item Reporting the experimental performance of our AIP approach for solving TDD programming tasks.
      \item Dissemination of the obtained theoretical and empirical results by submitting papers to top international conferences and journals in AI.        
      \end{enumerate}
\end{small}        
{\small {\bf Deliverable:} Final report with the obtained conclusions and produced publications.}
\end{enumerate}

Tasks ({\bf T1-T3}) will be developed by the three mentioned members of the research group: Sergio Jiménez, Eva Onaindia and Diego Aineto. In addition, we plan to hire a master student to assist in the completion of the task {\bf T3(a)}.

@@

Figure~\ref{fig:list} shows a synthesized program for traversing linked lists. The program is pictured as a {\em finite state machine}: the machine nodes mount to the different program lines while edges are tagged with a {\em condition/instruction} label, that denotes the condition (over the program variables) under which program instructions are taken.

@@

The objective of this project is to study the performance of our approach for the automated synthesis and validation of programs that are characterized by these tree dimensions: number of {\em program lines}, number of {\em program variables} and size of the available {\em instruction set}. In more detail, the two specific objectives of this project are:
\begin{itemize}
\item To study the performance of an Artificial Intelligence approach for the {\bf synthesis of programs} up to: {\bf 10 program lines}, {\bf 10 variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\item To study the performance of an Artificial Intelligence approach for the {\bf validation of programs} up to: {\bf 15 program lines}, {\bf 20 variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\end{itemize}
In both cases the performance of our approach will be evaluated with regard to the {\bf computation time} and {\bf memory} invested in the synthesize and the validation of the aimed programs.

Despite setting bounds for the programs size and kind, challenging programming tasks can still be addressed using {\em problem decomposition}. With this regard, our AIP encoding already supports callable procedures to decompose a given programming task in simpler modules and to enable recursive solutions~\cite{sergio:aprograming:icaps16,sergio:aprograming:ijcai16}.

\newpage
\section{Beneficios del proyecto. Difusión y Explotación en su caso de los Resultados.}
\label{subsec:beneficios}
Promising research opportunities come from the application of AIP to {\em program synthesis} given that, {\em program synthesis} with a tests base, can also be seen as a dual to {\em program testing}. AIP has recently shown successful in {\em program testing} to generate {\em attack plans} that completed non-trivial software security tests~\cite{hoffmann2015simulated,steinmetz2016revisiting,shmaryahu2016constructing,steinmetz2016goal}.

@@

The expected benefits of this project are:
\begin{enumerate}
\item The definition of {\bf effective metrics for assessing how well a given program covers a set of {\em input-output} test cases}.
\item An empirical {\bf study on the performance of the state-of-the-art AI planners for the {\em Synthesis and Validation of Programs}}.
\item The development, evaluation and dissemination of {\bf open source software for the {\em Synthesis and Validation of Programs}}.
\end{enumerate}

@@

The scientific results obtained during the development of the project will be submitted to top Artificial Intelligence conferences (such as IJCAI, AAAI, ICML and ICAPS) and to the main journals in the AI field (such as AIJ, JMLR and JAIR).

>>>>

This research project aims to investigate a novel approach for software development, that leverages Artificial Intelligence, to reduce the chances of introducing programming errors. The project proposes to address {\bf automatic program synthesis and validation starting from {\em input-output} tests cases, and using {\em AI planning} as a problem solving engine}. Our current work on this particular research topic is the recipient of the {\em 2016 distinguished paper award} at {\sc IJCAI}, the main international conference on Artificial Intelligence.
\end{abstract}

@@

\underline{General quality indicators of scientific research}: {\bf 373 citations} (287 since 2013) and {\bf h-index 11}. {\scriptsize\em Source google scholar}.

@@

Eva Onaindia is a {\em Full Professor} at the Universitat Politècnica of València. She leads the GRPS where she conducts research in Artificial Intelligence, Automated Planning, Causal Model Learning and Recommender Systems.

Eva has published numerous articles in top-tier specialized conferences such as the International Joint Conference on Artificial Intelligence (IJCAI), Autonomous Agents and Multi Agent Systems (AAMAS), International Conference on Planning and Scheduling (ICAPS) or European Conference on Artificial Intelligence (ECAI), among others; and scientific journals related to topics of Artificial Intelligence, Planning, Argumentation and Recommender Systems such as IEEE Intelligent Systems, Information Sciences, Engineering Applications of Artificial Intelligence, Knowledge and Information Systems or Applied Intelligence, among others.

She has led several National Research Projects as well as served on multiple ICAPS,AAAI, IJCAI and ECAI conferences as a member of the Program Committee and as a senior member. Eva is Editor-in-Chief of the journal AI Communications, a Review Board member of the journal Applied Intelligence and she will be Program co-chair of ICAPS 2019 (the main international conference on AI planning).

@@

\underline{General quality indicators of scientific research}: {\bf 1521 citations} (977 since 2013) and {\bf h-index 22}. {\scriptsize\em Source google scholar}.

@@

\item {\bf Test driven development (TDD)}~\cite{beck:TDD:2003} is a popular paradigm for software development that is frequently used in {\it agile methodologies}~\cite{cohen2003agile}. In TDD, test cases are created before the program code is written and they are run against the code during the development, e.g. after a code change via an automated process. When all tests pass, the program code is considered {\em complete} while when a test fails, it pinpoints a {\em bug} that must be fixed from the program code. Tests cases are a natural form of program specification, programmers often claim {\em 'code that is difficult to test is poorly written'}. Further, tests alert programmers of bugs before handing the code off to clients (the cost of finding a bug when the code is first written is considerably lower than the cost of detecting and fixing it later). Last but not least, writing a thorough set of tests cases forces programmers to think through inputs, outputs, and error conditions of programs. 
\end{itemize}

@@

Closely related to the aims of the project is also the work on the synthesis and validation of {\it Finite State Controllers} (FSCs)~\cite{geffner:policies:IJCAI15}. The state-of-the-art algorithms for computing FSCs follow a {\it top-down} approach that interleaves {\it programming} the FSC with validating it~\cite{sergio:aprograming:ijcai16}. To keep the computation of FSCs tractable, they limit the space of possible solutions bounding the maximum size of the FSC. The computation of FSCs includes works that compile this task into another forms of problem solving so they benefit from the last advances on off-the-shelf solvers (e.g. {\em classical planning}~\cite{sergio:aprograming:icaps16}, {\em conformant planning}~\cite{Geffner:FSM:AAAI10}, {\em CSP}~\cite{Infantes:FSC:ECAI2010} or a {\em Prolog program}~\cite{Giacomo:FSM:ICAPS13}).

Finally the generation of programs from examples is a research question also addressed in the classic AI field of {\em Inductive Logic Programming} (ILP)~\cite{muggleton1991inductive,Raedt:relationalML:book2008}. ILP arises from the intersection of {\em Machine Learning} and {\em Logic Programming} and deals with the development of inductive techniques to learn a given target concept, expressed as a  logic program,  from  examples  and  background  knowledge, that are expressed as logic facts.
\\
\\
\begin{center}
{\small (Please find the full list of {\em references} at the end of the document)}
\end{center}

@@

\subsection{Synthesis and validation of TDD programs as AI planning}
Given a {\em TDD programming} task our approach is modeling and solving it as it were an {\em AI planning} task.

Briefly the state variables of this AIP task are {\em fluents} of the kind:
\begin{itemize}
\item {\tt var:=value}, that encode the program variables).
\item {\tt program(line):=instruction}, that encode the program lines.
\item {\tt pcounter:=line}, encoding the current program line.
\end {itemize}

The initial/goal states of the AIP task encode the initial/final values of the program variables according to the input/output tests of the TDD programming task.

Finally, program instructions are encoded using two kinds of AIP actions:
\begin{itemize}
\item {\it Programming actions}, that set an instruction on a given program line.
\item {\it Execution actions}, execute the instruction set on a given program line.
\end{itemize}
We implement this encoding using standard planning languages, such as PDDL~\cite{fox2003pddl2}, so the AIP tasks resulting from our encoding can be solved with off-the-shelf planners, like {\sc FD}~\cite{helmert2006fast}. The programs synthesized with this approach are guaranteed to be bug-free over given sets of {\em input-output} tests cases.

@@

\subsection{Evaluation}
The performance of our approach for the synthesis and validation of TDD programs will be evaluated in two kinds of programming tasks:
\begin{itemize}
\item {\em Theoretical benchmarks}: Programming tasks for the computation of mathematical/logic series, string manipulation and for the management of classic data structures such as {\em lists}, {\em queues}, {\em stacks}, {\em trees}, {\em heaps}, \ldots
\item {\em Real-world benchmarks}: Practical programming tasks coming from {\em activity recognition} and {\em recommendation} domains. The GRPS group is currently leading a four-year research project from the Spanish National plan in which {\em activity recognition} and {\em activity recommendation} are applied to different real-world domains such as {\em tourism}, {\em domotics}, {\em traffic control} and {\em robotics}. We plan to evaluate our approach in programming and validation tasks that result from these domains.
\end{itemize}

@@

\begin{figure}[hbt!]
\begin{ganttchart}[
  hgrid,
  group progress label node/.append style={below=3pt},
  canvas/.append style={label=below:} ]{1}{12} 
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Design (6 months)}{1}{3} \\
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Development (6 months)}{4}{6}\\
\ganttbar[bar/.append style={line width=1pt, draw=TableBlue,fill=TableBlue,fill opacity=0.6274509804}]{Dissemination (12 months)}{7}{12}
\end{ganttchart}
\caption{\small Workplan for developing a user-interactive program synthesizer/validator.}
\label{fig:gantt}
\end{figure}

@@

{\small{\bf\em  Deliverable:} Technical report with the specifications of the system design.}

  \item {\bf T2. Development of the system architecture (months 7-12})
    \begin{small}
      \begin{enumerate}
      \item The programming-into-planning compiler ({\em Compiler 1}). This component parses the {\em TDD programming task} and produces an {\em AIP task} encoded in the standard planning language PDDL.
      \item The plan-into-program compiler ({\em Compiler 2}). This component extracts the program code and the corresponding validation certificate from the solution plan produced by an off-the-shelf AI planner.
      \end{enumerate}
\end{small}

@@

{\small{\bf\em Deliverable:} Open repository with the source code of the system architecture.}

\item {\bf T3. Experiments and dissemination of results (months 13-24}).
   \begin{small}
      \begin{enumerate}
      \item Reporting the experimental performance of our AIP approach for solving TDD programming tasks.
      \item Dissemination of the obtained theoretical and empirical results by submitting papers to top international conferences and journals in AI.        
      \end{enumerate}
\end{small}        
{\small{\bf\em  Deliverable:} Final report with the obtained conclusions and produced publications.}
\end{enumerate}

Tasks ({\bf T1-T3}) will be developed by the three mentioned members of the research group: Sergio Jiménez, Eva Onaindia and Diego Aineto. In addition, we plan to hire a master student to assist in the completion of task {\bf T3(a)}.

@@

Figure~\ref{fig:list} shows a synthesized program for traversing linked lists. The program is pictured as a {\em finite state machine}: The machine nodes mount to the different program lines while edges are tagged with a {\em condition/instruction} label, that denotes the condition (over the program variables) under which program instructions are taken.

@@

The objective of this project is to study the performance of our approach for the automated synthesis and validation of programs that are characterized by these tree dimensions: number of {\em program lines}, number of {\em program variables} and size of the available {\em instruction set}. For instance, the program of Figure~\ref{fig:list} is generated using three program lines, one observable variable $\mathsf{isEnd}(n)$ with Boolean domain, and a instruction set that comprises two instructions ($\mathsf{visit}(n)$ that marks the list node assigned to $n$ as {\em visited} and $\mathsf{step}(n)$ that advances $n$ to the next node in the linked list).

In more detail, the two specific objectives of this project are:
\begin{itemize}
\item To study the performance of our Artificial Intelligence approach for the {\bf synthesis of programs} up to: {\bf 10 program lines}, {\bf 10 variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\item To study the performance of our Artificial Intelligence approach for the {\bf validation of programs} up to: {\bf 15 program lines}, {\bf 20 variables} with binary domain and, an instruction set that comprises {\bf 10 instructions}. 
\end{itemize}
In both cases the performance of our approach will be evaluated with regard to the {\bf computation time} and {\bf memory} invested in the synthesize and the validation of the aimed programs.

Despite setting bounds for the programs size and kind, note that challenging programming tasks can be addressed using {\em problem decomposition}. With this regard, our AIP encoding already supports callable procedures to decompose a given programming task into simpler modules and to enable recursive solutions~\cite{sergio:aprograming:icaps16,sergio:aprograming:ijcai16}.

\newpage
\section{Beneficios del proyecto. Difusión y Explotación en su caso de los Resultados.}
\label{subsec:beneficios}
AIP has recently shown successful in {\em program testing} to generate {\em attack plans} that completed non-trivial software security tests~\cite{hoffmann2015simulated,steinmetz2016revisiting,shmaryahu2016constructing,steinmetz2016goal}. Promising research opportunities come from the application of AIP to {\em program synthesis} given that, {\em program synthesis} with a tests base, can also be seen as a dual to {\em program testing}.

@@

The expected benefits of this project are three-fold:
\begin{enumerate}
\item The definition of {\bf effective metrics for assessing how well a given program covers a set of {\em input-output} test cases}.
\item An empirical {\bf study on the performance of the state-of-the-art AI planners for the {\em Synthesis and Validation of Programs}}.
\item The development, evaluation and dissemination of {\bf open source software for the {\em Synthesis and Validation of Programs}}.
\end{enumerate}

@@

The scientific results obtained during the development of the project will also be submitted to top Artificial Intelligence conferences (such as IJCAI, AAAI, ICML and ICAPS) and to the main journals in the AI field (such as AIJ, JMLR and JAIR).

\vspace{0.5cm}
